\pkg{casebase}: An Alternative Framework For Survival Analysis

### 2019-12-11 analysis/paper/paper.tex
Bib+Why 60->14+packageCapitalization+cs3
fixed some bib issues, put a short description of why we went from 60 to 14, fixed package name capitalization issues and made case study 3 work with the attached support data.

\Address{
    Sahir Bhatnagar *\\
  McGill University\\
  1020 Pine Avenue West Montreal, QC, Canada H3A 1A2\\
  E-mail: \email{sahir.bhatnagar@mail.mcgill.ca}\\
  URL: \url{http://sahirbhatnagar.com/}\\~\\
      Maxime Turgeon *\\
  University of Manitoba\\
  186 Dysart Road Winnipeg, MB, Canada R3T 2N2\\
  E-mail: \email{max.turgeon@umanitoba.ca}\\
  URL: \url{https://maxturgeon.ca/}\\~\\
      Jesse Islam\\
  McGill University\\
  1020 Pine Avenue West Montreal, QC, Canada H3A 1A2\\
  E-mail: \email{jesse.islam@mail.mcgill.ca}\\

      James Hanley\\
  McGill University\\
  1020 Pine Avenue West Montreal, QC, Canada H3A 1A2\\
  E-mail: \email{james.hanley@mcgill.ca}\\
  URL: \url{http://www.medicine.mcgill.ca/epidemiology/hanley/}\\~\\
      Olli Saarela\\
  University of Toronto\\
  Dalla Lana School of Public Health, 155 College Street, 6th floor,
  Toronto, Ontario M5T 3M7, Canada\\
  E-mail: \email{olli.saarela@utoronto.ca}\\
  URL: \url{http://individual.utoronto.ca/osaarela/}\\~\\
  }

@@

@@

The purpose of the \pkg{casebase} package is to provide practitioners
with an easy-to-use software tool to predict the risk (or cumulative
incidence (CI)) of an event, for a particular patient. The following
points should be noted:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  hazard ratios
\item
  If, however, the absolute risks are of interest, they have to be
  recovered using the semi-parametric Breslow estimator
\item
  Alternative approaches for fitting flexible hazard models for
  estimating absolute risks, not requiring this two-step approach? Yes!
  \citep{hanley2009fitting}
\end{enumerate}

@@

These packages can belong to three main categories: non-proportional
hazards, regularization and competing risks. Besides these three main
categories, there are extra features of interest as well. Table
\ref{tab:surv-pkgs} summarizes the findings.

Many packages exist that implement semi-parametric Cox models. Of those,
\pkg{glmnet} \citeyearpar{regpathcox}???, \pkg{glmpath}
\citeyearpar{park_hastie}, \pkg{penalized} \citeyearpar{l1penal},
\pkg{RiskRegression} \citeyearpar{gerds_blanche} introduce penalization
in their model fitting. \pkg{CoxRidge} \citeyearpar{perperoglou},
\pkg{rstpm2} \citeyearpar{clements_liu} and \pkg{survival}
\citeyearpar{survival-package} include penalization, but they implement
the extended cox model permitting intuitive non proportional hazard
modeling while the previous four do not. In terms of penalization, the
benefits of \pkg{casebase} comes from its flexibility. Linear models
have a wide tool set, considerably larger than those made for Cox
regression. \pkg{casebase} permits many regularization tools developed
for linear regression to be repurposed within a survival context.

Parametric models permit non-proportional hazards and are implemented in
a handful of packages. Of interest are \pkg{CFC}
\citeyearpar{sharabiani_mahani_2019}, \pkg{flexsurv}
\citeyearpar{flexsurv}, \pkg{SmoothHazard} \citeyearpar{smoothHazard},
\pkg{rsptm2} \citeyearpar{clements_liu} and \pkg{survival}
\citeyearpar{survival-package}. \pkg{SmoothHazard} is limited to Weibull
distributions \citeyearpar{smoothHazard}, whereas \pkg{flexsurv} and
\pkg{survival} contain implementations for users to supply any
distribution they would like \citeyearpar{survival-package},
\citeyearpar{flexsurv}. \pkg{flexsurv}, \pkg{smoothhazard}, \pkg{mets}
and \pkg{rstpm2} can use splines, permitting a flexible fit over time
\citeyearpar{flexsurv}, \citeyearpar{smoothHazard},
\citeyearpar{scheike2014estimating}, \citeyearpar{clements_liu}. Though
\pkg{casebase} is limited to three distributions at the moment, it also
uses splines allowing for distributions that may not be consistent over
survival time.

Of the methods mentioned so far, only \pkg{CFC}, \pkg{flexsurv},
\pkg{mets} and \pkg{survival} contain implementations for competing
risks \citeyearpar{sharabiani_mahani_2019}, \citeyearpar{flexsurv},
\citeyearpar{scheike2014estimating}, \citeyearpar{survival-package}. The
differentiating factor between these packages and \pkg{casebase} is that
\pkg{casebase} handles competing risks through multiple logistic
regression. \pkg{SmoothHazard}, \pkg{survival} and \pkg{Rstpm2} can
handle interval censoring, whereas \pkg{casebase} currently cannot.
Interval censoring is a simple feature to implement and is a potential
improvement to add to \pkg{casebase}.

Of all the packages listed in Table \ref{tab:surv-pkgs}, \pkg{CFC},
\pkg{Flexsurv}, \pkg{mets}, \pkg{RiskRegression}, \pkg{rstpm2} and
\pkg{survival} have implementations for cumulative incidence
\citeyearpar{sharabiani_mahani_2019}, \citeyearpar{flexsurv},
\citeyearpar{scheike2014estimating}, \citeyearpar{gerds_blanche},
\citeyearpar{clements_liu}, \citeyearpar{survival-package}. Since
\pkg{casebase} is parametric, the baseline hazard is known and requires
integration to retrieve the cumulative incidence. This is similar to the
approaches of \pkg{CFC} and \pkg{flexsurv}
\citeyearpar{sharabiani_mahani_2019}, \citeyearpar{flexsurv}, while the
other packages use semi-parametric approaches to retrieve the baseline
after fitting \citeyearpar{scheike2014estimating},
\citeyearpar{gerds_blanche}. \pkg{rstpm2} and \pkg{survival} use both
for their respective parametric and semi-parametric model
implementations \citeyearpar{clements_liu},
\citeyearpar{survival-package}.

\begin{table}[ht]

@@

We examined the Study to Understand Prognoses Preferences Outcomes and
Risks of Treatment (SUPPORT) dataset (REF from main site). the SUPPORT
dataset tracks death for individuals who are considered seriously ill
within a hospital. Imputation was conducted using the default methods in
the mice package in R, available on CRAN. Before imputation, there was a
total of 1000 individuals and 35 variables. After imputation, we have
474 observations where 67.5\% of individuals in the study experience the
event of interest. The SUPPORT dataset will be used to demonstrate
regularization within casebase (REF), while comparing their absolute
risk preditions for a new individual. A description of each variable can
be found in Table\textasciitilde{}\ref{tab:support1} and a breakdown of
each categorical variable in Table\textasciitilde{}\ref{tab:support2}.
The data was downloaded from the main author's website, and it is also
available as part of the casebase package.

\begin{table}[ht]

@@

\begin{table}[ht]
\centering
\begin{tabular}{cc}
\hline
\textbf{Variable} & \textbf{Levels}                      \\
\hline
sex      & female                                        \\
         & male                                          \\
dzgroup  & ARF/MOSF w/Sepsis                             \\
         & COPD                                          \\
         & CHF                                           \\
         & Cirrhosis                                     \\
         & Coma                                          \\
         & Colon Cancer                                  \\
         & Lung Cancer                                   \\
         & MOSF w/Malig                                  \\
dzclass  & ARF/MOSF                                      \\
         & COPD/CHF/Cirrhosis                            \\
         & Coma                                          \\
         & Cancer                                        \\
income   & under \$11k                                   \\
         & $11-$25k                                      \\
         & $25-$50k                                      \\
         & \textgreater{}\$50k                           \\
race     & white                                         \\
         & black                                         \\
         & asian                                         \\
         & other                                         \\
         & hispanic                                      \\
sfdm2    & no(M2 and SIP pres)                           \\
         & adl\textgreater{}=4 (\textgreater{}=5 if sur) \\
         & SIP\textgreater{}=30                          \\
         & Coma or Intub                                 \\
         & \textless{}2 mo. follow-up                   
\end{tabular}
\caption{A description of each level within each categorical variable in the dataset, taken from REF.}
\label{tab:support2}
\end{table}

Hospital death as a covariate was removed, as this is directly
informative of death. The glmnet package(REF) on CRAN was used to
introduce elastic net, lasso and ridge regressions to casebase. To
visualize the effect of imputation on incidence density, we can refer to
the pre-imputation population-time plot in Figure \ref{fig:ptPreI} and
post-imputation population time plot in Figure \ref{fig:ptPostI}. The
incidence density is consistent between imputed and complete datasets.

@@

\begin{CodeInput}
R> #demonstrating incidence density before imputation
R> plot(casebase::popTime(as.data.frame(support),time = "d.time",event = "death"))
\end{CodeInput}
\end{CodeChunk}

\begin{CodeInput}
R> #demonstrating incidence density after imputation
R> plot(casebase::popTime(as.data.frame(workingCompleteData),time = "d.time",event = "death"))
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> #Estimating the absolute risk curve using the newData parameter.
R> casebaseAbsolute=casebase::absoluteRisk(cb.Model,time = seq(1,max(completeData$d.time), 0.5),newdata = t(newData) ,
R>                               s="lambda.1se",method=c("numerical"))
\end{CodeInput}
\end{CodeChunk}

@@

\hypertarget{case-study-4stanford-heart-transplant-data}{%
\section{Case study 4--Stanford Heart Transplant
Data}\label{case-study-4stanford-heart-transplant-data}}

Although the previous case studies provide a broad overview of the
capabilities of \pkg{casebase}, they all have two properties in common:

\begin{itemize}
  \item All covariates are fixed, i.e. they do not change over time;
  \item The estimated hazard functions satisfy the "proportional hazard" assumption.
\end{itemize}

Case-base sampling has been used in the literature to study vaccination
safety, where the exposure period was defined as the week following
vaccination \citep{saarela2015case}. Hence, the main covariate of
interest, i.e.~exposure to the vaccine, was changing over time. In this
context, case-base sampling offers an efficient alternative to nested
case-control designs or self-matching.

The purpose of the next case study is to show how \pkg{casebase} can
also be used for survival analysis problems that do not share the
properties above. To this end, we will use the Stanford Heart Transplant
Data \citep[\citet{crowley1977covariance}]{clark1971cardiac}; we will
use the version available in the \pkg{survival} package.

Recall the setting of this study: patients were admitted to the Stanford
program after meeting with their physician and determining that they
were unlikely to respond to other forms of treatment. After enrollment,
the program searches for a suitable donor for the patient, which can
take anywhere between a few days to almost a year. We are interested in
the effect of a heart transplant on survival; therefore, the patient is
considered exposed only after the transplant has occured.

As before, we can look at the population-time plot for a graphical
summary of the event incidence. As we can see, most events occur early
during the follow-up period, and therefore we do not expect the hazard
to be constant.

\begin{CodeInput}
R> library(survival)
R> library(casebase)
R> 
R> stanford_popTime <- popTime(jasa, time = "futime", 
R+                             event = "fustat")
R> plot(stanford_popTime)
\end{CodeInput}

\begin{center}\includegraphics{../figures/stanford-poptime-1} \end{center}

Since the exposure is time-dependent, we need to manually define the
exposure variable \emph{after} case-base sampling and \emph{before}
fitting the hazard function. For this reason, we will use the
\texttt{sampleCaseBase} function directly.

\begin{CodeInput}
R> library(tidyverse)
R> library(lubridate)
R> 
R> cb_data <- sampleCaseBase(jasa, time = "futime", 
R+                           event = "fustat", ratio = 10)
\end{CodeInput}
\end{CodeChunk}

Next, we will compute the number of days from acceptance into the
program to transplant, and we use this variable to determine whether
each population-moment is exposed or not.

\begin{CodeInput}
R> # Define exposure variable
R> cb_data <- mutate(cb_data,
R+                   txtime = time_length(accept.dt %--% tx.date, 
R+                                        unit = "days"),
R+                   exposure = case_when(
R+                     is.na(txtime) ~ 0L,
R+                     txtime > futime ~ 0L,
R+                     txtime <= futime ~ 1L
R+                   ))
\end{CodeInput}
\end{CodeChunk}

Finally, we can fit the hazard using various linear predictors.

\begin{CodeInput}
R> library(splines)
R> # Fit several models
R> fit1 <- fitSmoothHazard(fustat ~ exposure,
R+                         data = cb_data, time = "futime")
R> fit2 <- fitSmoothHazard(fustat ~ exposure + futime,
R+                         data = cb_data, time = "futime")
R> fit3 <- fitSmoothHazard(fustat ~ exposure + bs(futime),
R+                         data = cb_data, time = "futime")
R> fit4 <- fitSmoothHazard(fustat ~ exposure*bs(futime),
R+                         data = cb_data, time = "futime")
\end{CodeInput}
\end{CodeChunk}

Note that the fourth model (i.e.~\texttt{fit4}) includes an interaction
term between exposure and follow-up time. In other words, this model no
longer exhibit proportional hazards. The evidence of non-proportionality
of hazards in the Stanford Heart Transplant data has been widely
discussed \citep{arjas1988graphical}.

We can then compare the goodness of fit of these four models using the
Akaike Information Criterion (AIC).

\begin{CodeInput}
R> # Compute AIC
R> c("Model1" = AIC(fit1),
R+   "Model2" = AIC(fit2),
R+   "Model3" = AIC(fit3),
R+   "Model4" = AIC(fit4))
\end{CodeInput}

\begin{CodeOutput}
#> Model1 Model2 Model3 Model4 
#>    798    759    737    708
\end{CodeOutput}
\end{CodeChunk}

As we can, the best fit is the fourth model. By visualizing the hazard
functions for both exposed and unexposed individuals, we can more
clearly see how the hazards are no longer proportional.

\begin{CodeInput}
R> # Compute hazards---
R> # First, create a list of time points for both exposure status
R> hazard_data <- expand.grid(exposure = c(0, 1),
R+                            futime = seq(0, 1000,
R+                                         length.out = 100))
R> # Set the offset to zero
R> hazard_data$offset <- 0 
R> # Use predict to get the fitted values, and exponentiate to 
R> # transform to the right scale
R> hazard_data$hazard = exp(predict(fit4, newdata = hazard_data,
R+                                  type = "link"))
R> # Add labels for plots
R> hazard_data$Status = factor(hazard_data$exposure,
R+                             labels = c("NoTrans", "Trans"))
R> 
R> ggplot(hazard_data, aes(futime, hazard, colour = Status)) +
R+     geom_line() +
R+     theme_minimal() +
R+     theme(legend.position = 'top') +
R+     ylab('Hazard') + xlab('Follow-up time')
\end{CodeInput}

\begin{center}\includegraphics{../figures/stanford-hazard-1} \end{center}

The non-proportionality seems to be more pronounced at the beginning of
follow-up than the end. Finally, we can turn these estimate of the
hazard function into estimates of the cumulative incidence functions.

\begin{CodeInput}
R> # Compute absolute risk curves
R> newdata <- data.frame(exposure = c(0, 1))
R> absrisk <- absoluteRisk(fit4, newdata = newdata, 
R+                         time = seq(0, 1000, length.out = 100))
R> 
R> colnames(absrisk) <- c("Time", "NoTrans", "Trans")
R> 
R> # Rearrange the data
R> absrisk <- gather(as.data.frame(absrisk),
R+                   "Status", "Risk", -Time)
R>  
R> ggplot(absrisk, aes(Time, Risk, colour = Status)) +
R+   geom_line() +
R+   theme_minimal() +
R+   theme(legend.position = 'top') +
R+   expand_limits(y = 0.5) +
R+   xlab('Follow-up time') + ylab('Cum. Incidence')
\end{CodeInput}

\begin{center}\includegraphics{../figures/stanford-risk-1} \end{center}

Note that we can easily adapt the code above to the situation where a
patient receives a heart transplant at a point in time of interest, for
example after 30 days.

\begin{CodeInput}
R> # Compute hazards---
R> # First, create a list of time points for both exposure status
R> one_yr_haz <- data.frame(futime = seq(0, 365,
R+                                       length.out = 100))
R> one_yr_haz <- mutate(one_yr_haz,
R+                      offset = 0,
R+                      exposure = if_else(futime < 30, 0, 1))
R> one_yr_haz$hazard = exp(predict(fit4, newdata = one_yr_haz,
R+                                 type = "link"))
R> ggplot(one_yr_haz, aes(futime, hazard)) +
R+   geom_line() +
R+   theme_minimal() +
R+   theme(legend.position = 'top') +
R+   ylab('Hazard') + xlab('Follow-up time') +
R+   geom_vline(xintercept = 30, linetype = 'dashed')
\end{CodeInput}

\begin{center}\includegraphics{../figures/stanford-1y-haz-1} \end{center}

We can then compare the 1-year mortality risk without transplant and
with transplant at 30 days.

\begin{CodeInput}
R> absoluteRisk(fit4, newdata = data.frame(exposure = 0),
R+              time = 365)
\end{CodeInput}

\begin{CodeOutput}
#>         
#> 365 0.32
\end{CodeOutput}

\begin{CodeInput}
R> 
R> # Use the trapezoidal rule to estimate the cumulative hazard
R> min_hazard <- min(one_yr_haz$hazard)
R> max_hazard <- max(one_yr_haz$hazard)
R> increment <- 365/99
R> cumhaz_est <- 0.5*increment*(sum(2*one_yr_haz$hazard) - min_hazard - max_hazard)
R> 1 - exp(-cumhaz_est)
\end{CodeInput}

\begin{CodeOutput}
#> [1] 0.24
\end{CodeOutput}
\end{CodeChunk}

As we can see, the risk estimate at 1-year is about 30\% lower if the
patient receives a heart transplant at 30 days.

@@

This report was generated on 2019-12-08 12:31:41 using the following
computational environment and dependencies:

@@

#> - Session info ----------------------------------------------------------

@@

#>  tz       America/Winnipeg            
#>  date     2019-12-08                  

@@

#> - Packages --------------------------------------------------------------
#>  package     * version    date       lib source                          
#>  assertthat    0.2.1      2019-03-21 [1] CRAN (R 3.6.1)                  
#>  backports     1.1.5      2019-10-02 [1] CRAN (R 3.6.1)                  
#>  broom         0.5.2      2019-04-07 [1] CRAN (R 3.6.1)                  
#>  callr         3.3.1      2019-07-18 [1] CRAN (R 3.6.1)                  
#>  casebase    * 0.2.1.9001 2019-07-29 [1] local                           
#>  cellranger    1.1.0      2016-07-27 [1] CRAN (R 3.5.0)                  
#>  cli           1.1.0      2019-03-19 [1] CRAN (R 3.6.1)                  
#>  codetools     0.2-16     2018-12-24 [4] CRAN (R 3.5.2)                  
#>  colorspace    1.4-1      2019-03-18 [1] CRAN (R 3.6.1)                  
#>  crayon        1.3.4      2017-09-16 [1] CRAN (R 3.5.0)                  
#>  data.table    1.12.2     2019-04-07 [1] CRAN (R 3.6.1)                  
#>  desc          1.2.0      2018-05-01 [1] CRAN (R 3.5.0)                  
#>  devtools      2.1.0      2019-07-06 [1] CRAN (R 3.6.1)                  
#>  digest        0.6.22     2019-10-21 [1] CRAN (R 3.6.1)                  
#>  dplyr       * 0.8.3      2019-07-04 [1] CRAN (R 3.6.1)                  
#>  ellipsis      0.3.0      2019-09-20 [1] CRAN (R 3.6.1)                  
#>  evaluate      0.14       2019-05-28 [1] CRAN (R 3.6.1)                  
#>  forcats     * 0.4.0      2019-02-17 [1] CRAN (R 3.6.1)                  
#>  foreach     * 1.4.4      2017-12-12 [1] CRAN (R 3.5.0)                  
#>  fs            1.3.1      2019-05-06 [1] CRAN (R 3.6.1)                  
#>  generics      0.0.2      2018-11-29 [1] CRAN (R 3.6.1)                  
#>  ggplot2     * 3.2.1      2019-08-10 [1] CRAN (R 3.6.1)                  
#>  glmnet      * 2.0-18     2019-05-20 [1] CRAN (R 3.6.1)                  
#>  glue          1.3.1      2019-03-12 [1] CRAN (R 3.6.1)                  
#>  gtable        0.3.0      2019-03-25 [1] CRAN (R 3.6.1)                  
#>  haven         2.1.1      2019-07-04 [1] CRAN (R 3.6.1)                  
#>  hms           0.5.2      2019-10-30 [1] CRAN (R 3.6.1)                  
#>  htmltools     0.3.6      2017-04-28 [1] CRAN (R 3.5.0)                  
#>  httr          1.4.1      2019-08-05 [1] CRAN (R 3.6.1)                  
#>  iterators     1.0.9      2017-12-12 [1] CRAN (R 3.5.0)                  
#>  jsonlite      1.6        2018-12-07 [1] CRAN (R 3.6.1)                  
#>  knitr         1.23       2019-05-18 [1] CRAN (R 3.6.1)                  
#>  labeling      0.3        2014-08-23 [1] CRAN (R 3.5.0)                  
#>  lattice       0.20-38    2018-11-04 [4] CRAN (R 3.5.1)                  
#>  lazyeval      0.2.2      2019-03-15 [1] CRAN (R 3.6.1)                  
#>  lifecycle     0.1.0      2019-08-01 [1] CRAN (R 3.6.1)                  
#>  lubridate   * 1.7.4      2018-04-11 [1] CRAN (R 3.5.0)                  
#>  magrittr    * 1.5        2014-11-22 [1] CRAN (R 3.5.0)                  
#>  Matrix      * 1.2-18     2019-11-27 [4] CRAN (R 3.6.1)                  
#>  memoise       1.1.0      2017-04-21 [1] CRAN (R 3.5.0)                  
#>  mgcv          1.8-24     2018-06-18 [1] CRAN (R 3.5.1)                  
#>  modelr        0.1.5      2019-08-08 [1] CRAN (R 3.6.1)                  
#>  munsell       0.5.0      2018-06-12 [1] CRAN (R 3.6.1)                  
#>  nlme          3.1-142    2019-11-07 [4] CRAN (R 3.6.1)                  
#>  pillar        1.4.2      2019-06-29 [1] CRAN (R 3.6.1)                  
#>  pkgbuild      1.0.3      2019-03-20 [1] CRAN (R 3.6.1)                  
#>  pkgconfig     2.0.3      2019-09-22 [1] CRAN (R 3.6.1)                  
#>  pkgload       1.0.2      2018-10-29 [1] CRAN (R 3.6.1)                  
#>  prettyunits   1.0.2      2015-07-13 [1] CRAN (R 3.5.2)                  

>>>>

\Address{
    Sahir Bhatnagar *\\
  McGill Univeristy\\
  1020 Pine Avenue West Montreal, QC, Canada H3A 1A2\\
  E-mail: \email{sahir.bhatnagar@mail.mcgill.ca}\\
  URL: \url{http://sahirbhatnagar.com/}\\~\\
      Maxime Turgeon *\\
  University of Manitoba\\
  318 Machray Hall Winnipeg, MB, Canada R3T 2N2\\
  E-mail: \email{max.turgeon@umanitoba.ca}\\
  URL: \url{http://maxturgeon.ca/}\\~\\
      Jesse Islam\\
  McGill University\\
  1020 Pine Avenue West Montreal, QC, Canada H3A 1A2\\
  E-mail: \email{jesse.islam@mail.mcgill.ca}\\

      James Hanley\\
  McGill Univeristy\\
  1020 Pine Avenue West Montreal, QC, Canada H3A 1A2\\
  E-mail: \email{james.hanley@mcgill.ca}\\
  URL: \url{http://www.medicine.mcgill.ca/epidemiology/hanley/}\\~\\
      Olli Saarela\\
  University of Toronto\\
  Dalla Lana School of Public Health, 155 College Street, 6th floor,
  Toronto, Ontario M5T 3M7, Canada\\
  E-mail: \email{olli.saarela@utoronto.ca}\\
  URL: \url{http://individual.utoronto.ca/osaarela/}\\~\\
  }

@@

\DeclareUnicodeCharacter{2500}{}

\usepackage{tabularx}

@@

The purpose of the \texttt{casebase} package is to provide practitioners
with an easy-to-use software tool to predict the risk (or cumulative
incidence (CI)) of an event, for a particular patient. The following
points should be noted:

hazard ratios 2. If, however, the absolute risks are of interest, they
have to be recovered using the semi-parametric Breslow estimator 3.
Alternative approaches for fitting flexible hazard models for estimating
absolute risks, not requiring this two-step approach? Yes!
\citep{hanley2009fitting}

@@

These packages can belong to three main categories: non-proportional
hazards, regularization and competing risks. Besides these three main
categories, there are extra features of interest as well. Table XXX
summarizes the findings.

Many packages exist that implement semi-parametric cox models. Of those,
glmnet \citeyearpar{regpathcox}???, glmpath \citeyearpar{park_hastie},
penalized \citeyearpar{l1penal}, RiskRegression
\citeyearpar{gerds_blanche} introduce penalization in their model
fitting. CoxRidge \citeyearpar{perperoglou} , rstpm2
\citeyearpar{clements_liu} and survival \citeyearpar{survival-package}
include penalization, but they implement the extended cox model
permitting intuitive non proportional hazard modeling while the previous
four do not. In terms of penalization, the benefit of casebase comes
from its flexibility. Linear models have a wide tool set, considerably
larger than those made for cox regression (\emph{reference for number of
packages between their task views, regression is spread out across
Multivariate, SocialSciences, Robust among others. The focus of this
task view is on Optimization Infrastructure Packages , General Purpose
Continuous Solvers , Mathematical Programming Solvers , and Specific
Applications in Optimization , or Multi Objective Optimization}).
Casebase permits many regularization tools developed for linear
regression to be repurposed within a survival context.

Parametric models permit non-proportional hazards and are implemented in
a handful of packages. Of interest are CFC
\citeyearpar{sharabiani_mahani_2019}, flexsurv \citeyearpar{flexsurv} ,
SmoothHazard \citeyearpar{smoothHazard}, Rsptm2
\citeyearpar{clements_liu} and Survival \citeyearpar{survival-package}.
Smooth hazard is limited to weibull
distributions\citeyearpar{smoothHazard}, where as flexsurv and survival
contain implementations for users to supply any distribution they would
like \citeyearpar{survival-package}, \citeyearpar{flexsurv}. Flexsurv,
smooth hazard, mets and rstpm2 can use splines, permitting a flexible
fit over time \citeyearpar{flexsurv}, \citeyearpar{smoothHazard},
\citeyearpar{scheike2014estimating}, \citeyearpar{clements_liu} . Though
casebase is limited to 3 distributions at the moment, it also uses
splines allowing for distributions that may not be consistent over
survival time.

Of the methods mentioned so far, only CFC,flexsurv, mets and survival
contain implementations for competing risks
\citeyearpar{sharabiani_mahani_2019},
\citeyearpar{flexsurv},\citeyearpar{scheike2014estimating}
,\citeyearpar{survival-package} (\emph{how is it done for parametric
models?}). The differentiating factor between these packages and
casebase is that casebase handles competing risks through multiple
logistic regression. Smooth hazard, survival and Rstpm2 can handle
interval censoring, where as casebase currently can not. Interval
censoring is simple feature to implement and is a potential improvement
to add to casebase. (There's also multistate models, which I haven't
mentioned in this paper)

Of all the packages listed in the table, CFC , Flexsurv, mets,
RiskRegression, rstpm2 and survival have implementations for cumulative
incidence \citeyearpar{sharabiani_mahani_2019}, \citeyearpar{flexsurv},
\citeyearpar{scheike2014estimating}, \citeyearpar{gerds_blanche},
\citeyearpar{clements_liu}, \citeyearpar{survival-package}. Since
casebase is parametric, the baseline hazard is known and requires
integration to retrieve the cumulative incidence. This is similar to the
approaches of CFC and flexsurv \citeyearpar{sharabiani_mahani_2019},
\citeyearpar{flexsurv}, while the other packages use semi-parametric
approaches to retrieve the baseline after fitting
\citeyearpar{scheike2014estimating}, \citeyearpar{gerds_blanche}. rstpm2
and survival use both for their respective parametric and
semi-parametric model implementations \citeyearpar{clements_liu},
\citeyearpar{survival-package}.

\begin{table}[]

@@

We examined the Study to Understand Prognoses Preferences Outcomes and
Risks of Treatment (SUPPORT) dataset (REF from main site). the SUPPORT
dataset tracks death for individuals who are considered seriously ill
within a hospital. Imputation was conducted using the default methods in
the mice package in R, available on CRAN. Before imputation, there was a
total of 1000 individuals and 35 variables. After imputation, we have
474 observations where 67.5\% of individuals in the study experience the
event of interest. The SUPPORT dataset will be used to demonstrate
regularization within casebase (REF), while comparing their absolute
risk preditions for a new individual. A description of each variable can
be found in table XXX1 and a breakdown of each categorical variable in
table XXX2. The data was downloaded from the main author's website, and
it is also available as part of the casebase package.

\begin{table}[]

@@

\begin{table}[]
\centering
\begin{tabular}{cc}
\hline
\textbf{Variable} & \textbf{Levels}                      \\
\hline
sex      & female                                        \\
         & male                                          \\
dzgroup  & ARF/MOSF w/Sepsis                             \\
         & COPD                                          \\
         & CHF                                           \\
         & Cirrhosis                                     \\
         & Coma                                          \\
         & Colon Cancer                                  \\
         & Lung Cancer                                   \\
         & MOSF w/Malig                                  \\
dzclass  & ARF/MOSF                                      \\
         & COPD/CHF/Cirrhosis                            \\
         & Coma                                          \\
         & Cancer                                        \\
income   & under \$11k                                   \\
         & $11-$25k                                      \\
         & $25-$50k                                      \\
         & \textgreater{}\$50k                           \\
race     & white                                         \\
         & black                                         \\
         & asian                                         \\
         & other                                         \\
         & hispanic                                      \\
sfdm2    & no(M2 and SIP pres)                           \\
         & adl\textgreater{}=4 (\textgreater{}=5 if sur) \\
         & SIP\textgreater{}=30                          \\
         & Coma or Intub                                 \\
         & \textless{}2 mo. follow-up                   
\end{tabular}
\caption{A description of each level within each categorical variable in the dataset, taken from REF.}
\label{tab:my-table}
\end{table}

Hospital death as a covariate was removed, as this is directly
informative of death. The glmnet package(REF) on CRAN was used to
introduce elastic net, lasso and ridge regressions to casebase. To
visualize the incidence density, we can refer to the post-imputation
population-time plot in Figure \ref{fig:ptPostI}.

@@

\begin{CodeInput}
R> #demonstrating incidence density after imputation
R> plot(casebase::popTime(as.data.frame(completeData),time = "d.time",event = "death"))
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> #Estimating the absolute risk curve using the newData parameter.
R> casebaseAbsolute=casebase::absoluteRisk(cb.Model,time = seq(1,max(completeData$d.time), 0.5),newdata = t(newData) , s="lambda.1se",method=c("numerical"))
\end{CodeInput}
\end{CodeChunk}

@@

@@

This report was generated on 2019-12-11 12:28:29 using the following
computational environment and dependencies:

@@

#> ─ Session info ───────────────────────────────────────────────────────────────

@@

#>  tz       America/Montreal            
#>  date     2019-12-11                  

@@

#> ─ Packages ───────────────────────────────────────────────────────────────────
#>  package     * version    date       lib
#>  assertthat    0.2.1      2019-03-21 [1]
#>  backports     1.1.5      2019-10-02 [1]
#>  broom         0.5.2      2019-04-07 [1]
#>  callr         3.4.0      2019-12-09 [1]
#>  casebase    * 0.2.1.9001 2019-12-11 [1]
#>  cellranger    1.1.0      2016-07-27 [1]
#>  cli           2.0.0      2019-12-09 [1]
#>  codetools     0.2-16     2018-12-24 [1]
#>  colorspace    1.4-1      2019-03-18 [1]
#>  crayon        1.3.4      2017-09-16 [1]
#>  data.table    1.12.8     2019-12-09 [1]
#>  DBI           1.0.0      2018-05-02 [1]
#>  dbplyr        1.4.2      2019-06-17 [1]
#>  desc          1.2.0      2018-05-01 [1]
#>  devtools      2.2.1      2019-09-24 [1]
#>  digest        0.6.23     2019-11-23 [1]
#>  dplyr       * 0.8.3      2019-07-04 [1]
#>  ellipsis      0.3.0      2019-09-20 [1]
#>  evaluate      0.14       2019-05-28 [1]
#>  fansi         0.4.0      2018-10-05 [1]
#>  forcats     * 0.4.0      2019-02-17 [1]
#>  foreach       1.4.7      2019-07-27 [1]
#>  fs            1.3.1      2019-05-06 [1]
#>  generics      0.0.2      2018-11-29 [1]
#>  ggplot2     * 3.2.1      2019-08-10 [1]
#>  glmnet      * 3.0-1      2019-11-15 [1]
#>  glue          1.3.1      2019-03-12 [1]
#>  gtable        0.3.0      2019-03-25 [1]
#>  haven         2.2.0      2019-11-08 [1]
#>  hms           0.5.2      2019-10-30 [1]
#>  htmltools     0.4.0      2019-10-04 [1]
#>  httr          1.4.1      2019-08-05 [1]
#>  iterators     1.0.12     2019-07-26 [1]
#>  jsonlite      1.6        2018-12-07 [1]
#>  knitr         1.26       2019-11-12 [1]
#>  lattice       0.20-38    2018-11-04 [1]
#>  lazyeval      0.2.2      2019-03-15 [1]
#>  lifecycle     0.1.0      2019-08-01 [1]
#>  lubridate     1.7.4      2018-04-11 [1]
#>  magrittr    * 1.5        2014-11-22 [1]
#>  Matrix      * 1.2-17     2019-03-22 [1]
#>  memoise       1.1.0      2017-04-21 [1]
#>  mgcv          1.8-31     2019-11-09 [1]
#>  modelr        0.1.5      2019-08-08 [1]
#>  munsell       0.5.0      2018-06-12 [1]
#>  nlme          3.1-142    2019-11-07 [1]
#>  pillar        1.4.2      2019-06-29 [1]
#>  pkgbuild      1.0.6      2019-10-09 [1]
#>  pkgconfig     2.0.3      2019-09-22 [1]
#>  pkgload       1.0.2      2018-10-29 [1]
#>  prettyunits   1.0.2      2015-07-13 [1]
#>  processx      3.4.1      2019-07-18 [1]
#>  ps            1.3.0      2018-12-21 [1]
#>  purrr       * 0.3.3      2019-10-18 [1]
#>  R6            2.4.1      2019-11-12 [1]
#>  Rcpp          1.0.3      2019-11-08 [1]
#>  readr       * 1.3.1      2018-12-21 [1]
#>  readxl        1.3.1      2019-03-13 [1]
#>  remotes       2.1.0      2019-06-24 [1]
#>  reprex        0.3.0      2019-05-16 [1]
#>  rlang         0.4.2      2019-11-23 [1]
#>  rmarkdown     1.18       2019-11-27 [1]
#>  rprojroot     1.3-2      2018-01-03 [1]
#>  rstudioapi    0.10       2019-03-19 [1]
#>  rticles       0.12       2019-11-12 [1]
#>  rvest         0.3.5      2019-11-08 [1]
#>  scales        1.1.0      2019-11-18 [1]
#>  sessioninfo   1.1.1      2018-11-05 [1]
#>  shape         1.4.4      2018-02-07 [1]
#>  stringi       1.4.3      2019-03-12 [1]
#>  stringr     * 1.4.0      2019-02-10 [1]
#>  survival    * 3.1-7      2019-11-09 [1]
#>  testthat      2.3.1      2019-12-01 [1]
#>  tibble      * 2.1.3      2019-06-06 [1]
#>  tidyr       * 1.0.0      2019-09-11 [1]
#>  tidyselect    0.2.5      2018-10-11 [1]
#>  tidyverse   * 1.3.0      2019-11-21 [1]
#>  usethis       1.5.1      2019-07-04 [1]
